Compositional equivalences based on Open pNets

### 2021-10-22 2016-Weak-FH-Bisim/WeakBisim.tex
typos and lmcs style

\subsubsection*{Structure}
This article is organised as follows. Section~\ref{sec:notations}
provides the definition of pNets and introduces the notations used in
this paper, including the definition of open pNets. 
Section~\ref{sec:OT} defines open automata, i.e. automata
with parameters and transitions conditioned by the behaviour of
``holes''; a strong bisimulation equivalence for open automata is also
presented in this section. 
Section~\ref{section:op-semantics} gives
the semantics of open pNets expressed as open automata, and states
compositional properties on the strong bisimulation for open
pNets. 
Section~\ref{sec:weak} defines a weak bisimulation
equivalence on open automata and derives weak bisimilarity for pNets,
together with properties on compositionality of weak bisimulaiton for
open pNets. 
Finally, Section~\ref{sec:RW} discusses related works and Section~\ref{section:conclusion} concludes the
paper.

@@

The informal semantics of pNets is as follows. pLTSs behave more or less like  classical automata with conditional branching and variables. The actions on the \pLTS s can send or receive values, potentially modifying the value of variables. 
pNets are synchronisation entities: a pNet node composes several sub-pNets and  define how the sub-pNets interact, where a sub-pNet is either a pNet or a pLTS. The synchronisation between sub-pNets is defined by synchronisation vectors (originally introduced by \cite{Arnold1982}) that express how an action of a sub-pNet can be synchronised with actions of other sub-pNet, and how the resulting synchronised action is visible from outside of the pNet. The synchronisation mechanism is very expressive, including pattern-matching/unification between the parameterized actions within the vector, and an additional predicate over their variables.
Consider a pNet node that assembles several pLTSs, the synchronisation vectors specify the way that transitions of the composed pNet are built from the transitions of the sub-nets. This can be seen as "conditional transtions" of the pNet, or alternatively, as a syntax to encode structural operational semantics (SOS rules) of the system: each vector expresses not only the actions emitted by the pNet but also what transitions of the composed pLTSs must occur to trigger this global transition.
Synchronisation vectors can also express the exportation of an action of a sub-pNet to the next level, or to hide an interaction and make it non-observable. Finally, a pNet can leave sub-pNets undefined and instead declare holes with a well-defined signature. Holes can then be filled with a sub-pNet. This is defined as follows.

@@

Figure \ref{SimpleProt:Spec} (right) shows a graphical view of the pNet $\symb{SimpleProtocolSpec}$ that specifies the system. The pNet is made of the composition of two pNets: a $\symb{SimpleSystem}$ node, and a $\symb{PerfectBuffer}$ sub-pNet. 
The full system implementation should be equivalent (e.g.  weakly bisimilar)
to thhis $\symb{SimpleProtocolSpec}$.  
The pNet has a tree-like structure.  The root node of the tree {\it SimpleSystem} is the top level of the pNet structure. It acts as the parallel operator. It consists of three nodes: two holes $P$ and $Q$ and one sub-pNet, denoted {\it PerfectBuffer}. Nodes of the tree are synchronised using four synchronisation vectors, that  express the possible synchronisations between the parameterised actions of a subset of the nodes. For instance, in the vector $< \texttt{p-send(m)},\texttt{in(m)},\_> \rightarrow \texttt{\nounderline{in(m)}}$ only $P$ and {\it PerfectBuffer} nodes are involved in the synchronisation. The synchronisation between these processes occurs when  process $P$  performs $\texttt{p-send(m)}$  action sending a message, and  the {\it PerfectBuffer} accepts the message through an $\texttt{in(m)}$ action  at the same time; the result that will be returned at upper level is the action  $\texttt{\nounderline{in(m)}}$.

@@

\begin{theorem}[FH-Bisimulation is an equivalence]\label{thm-equiv} Suppose $\mathcal{R}$ 
is an FH-bisimualtion. Then $\mathcal{R}$ is an equivalence, that is, $\mathcal{R}$ is 
reflexive, symmetric and transitive.
\end{theorem}

@@

\begin{theorem}[Decidability of FH-bisimulation]
Let $A_1$ and $A_2$ be finite open automata
and $\mathcal{R}$ a relation over their states $\mathcal{S}_1$ and
$\mathcal{S}_2$ constrained by a set of predicates. Assume that
the predicates inclusion is decidable over  
the action algebra $\AlgA$. Then it is decidable whether the relation 
$\mathcal{R}$ is an FH-bisimulation.

\end{theorem}

@@

We start with a decomposition property: from one open transition of $P[Q]_{j_0}$, we exhibit 
corresponding behaviours of $P$ and $Q$, and determine the relation between their 
predicates.
\begin{lemma}[Open transition decomposition\label{lem-decompose}] Consider two pNets $P$ and $Q$ that are not pLTSs\footnote{A similar lemma can be proven for a pLTS $Q$}.
	Let $\Leaves(Q)=p_l^{l\in L_Q}$ and suppose:
	\[ P[Q]_{j_0}  
		\models
		{\openrule
			{
				\beta_j^{j\in J}, \Pred,  
				\Post}
			{\ostate{s_i^{i\in L}} \OTarrow {\alpha}
				\ostate{s_i'^{\, i\in L}}}
		}
	\]
		with  $J\cap\Holes(Q)\neq\emptyset$ or $\exists i\in L_Q.\,s_i\neq s'_i$, i.e. $Q$ takes part in the reduction.
		 Then there exist $\alpha_Q$, $\Pred\,'$, $\Pred\,''$, 
		$\Post\,'$, $\Post\,''$ s.t.:\\[-2ex]
		%\[
		\begin{mathpar}
		P\models{\openrule
			{
				\beta_j^{j\in (J\setminus \Holes(Q)) \cup \{j_0\}}, 
				\Pred\,',  
				\Post\,'}
			{\ostate{s_i^{i\in L\setminus L_Q}} \OTarrow {\alpha}
				\ostate{s_i'^{\,i\in L\setminus L_Q}}}
		}%\]
	\vspace{-2.2ex}\\\text{and~\qquad\qquad~}
		%\[
		Q\models{\openrule
			{
				\beta_j^{j\in J\cap\Holes(Q)}, \Pred\,'',  
				\Post\,''}
			{\ostate{s_i^{i\in L_Q}} \OTarrow {\alpha_Q}
				\ostate{s_i'^{\,i\in L_Q}}}
		}%\]
		\end{mathpar}
		and  $\Pred \iff \Pred\,'
		\land \Pred\,''\land \alpha_Q=\beta_{j_0}$, $\Post=\Post\,'\uplus 
		\Post\,''$ where $\Post\,''$ is the restriction of $\Post$ over variables of 
		$Q$.
\end{lemma}

@@

\begin{lemma}[Open transition composition]\label{lem-compose} 
	Suppose $j_0\in J$ and:\\[-1ex]
\begin{mathpar}
%\[
P\models{\openrule
	{
		\beta_j^{j\in J}, 
		\Pred,  
		\Post}
	{\ostate{s_i^{i\in L}} \OTarrow {\alpha}
		\ostate{s_i'^{\, i\in L}}}
}%\]
\quad\text{~~and~~}\quad
%\[
Q\models{\openrule
	{
		\beta_j^{j\in J_Q},
		 \Pred\,',  
		\Post\,'}
	{\ostate{s_i^{i\in L_Q}} \OTarrow {\alpha_Q}
		\ostate{s_i'^{\, i\in L_Q}}}
}%\]
\end{mathpar}
Then, we have:\\[-1ex]
	\[ P[Q]_{j_0}  
	\models
	{\openrule
		{
			\beta_j^{(j\in J\setminus\{j_0\}) \uplus J_Q}, 
			\Pred\land\Pred\,'\land \alpha_Q=\beta_{j_0},  
			\Post\uplus \Post\,'}
		{\ostate{s_i^{i\in L\uplus L_Q}} \OTarrow {\alpha}
			\ostate{s_i'^{\, i\in L\uplus L_Q}}}
	}
	\]
\end{lemma}

@@

\begin{theorem}[Congruence]\label{thm-congr-eq}
	Consider an open pNet:
	$\pNet = \mylangle \pNet_i^{i\in I}, \Sort_j^{j\in J}, 
	\set{\symb{SV}}\myrangle$.
	Let $j_0\in J$ be a hole. Let $\pNetQ$ and $\pNetQ'$ be two FH-bisimilar pNets such that\footnote{Note that $\Sortop(\pNetQ)=\Sortop(\pNetQ')$ is 
	ensured by 
	strong bisimilarity.} 
	$\Sortop(\pNetQ)=\Sortop(\pNetQ')=\Sort_{j_0}$. Then 
	$\pNet[\pNetQ]_{j_0}$ and 
	$\pNet[\pNetQ']_{j_0}$ are FH-bisimilar.
\end{theorem}

\begin{theorem}[Context equivalence]\label{thm-ctxt-eq}
	Consider two  open pNets
	$\pNet = \mylangle \pNet_i^{i\in I}, \Sort_j^{j\in J}, 
	\set{\symb{SV}}\myrangle$ and 	$\pNet' = \mylangle {\pNet'}_i^{i\in I}, 
	\Sort_j^{j\in 
	J}, 	\set{\symb{SV'}}\myrangle$ that are FH-bisimilar
	(recall they must have the same holes to be bisimilar).
	Let $j_0\in J$ be a hole, and $Q$ be a pNet such that $\Sortop(Q)=\Sort_{j_0}$. Then 
	$\pNet[Q]_{j_0}$ and 
	$\pNet'[Q]_{j_0}$ are FH-bisimilar.
\end{theorem}

Finally, the previous theorems can be composed to state a general theorem about 
composability and FH-bisimilarity.
\begin{theorem}[Composability] \label{thm-composability}
	Consider two FH-bisimilar pNets with an arbitrary number of holes, when replacing, 
	inside those two original pNets, a subset of the holes by FH-bisimilar pNets, we 
	obtain two FH-bisimilar pNets.
\end{theorem}
This theorem is quite powerful. It somehow implies that the theory of open pNets is convenient to study properties of process composition. Open pNets can indeed be used to study process operators and process algebras, as shown in~\cite{henrio:Forte2016} where compositional properties are extremely useful. In the case of interaction protocols~\cite{BHHM:FACS11}, composition of bisimulation can justify abstractions used in some parts of the application.

@@

\begin{theorem}[Weak FH-Bisimulation is an equivalence]\label{thm-weak-equiv} Suppose $\mathcal{R}$ 
is a weak FH-bisimulation. Then $\mathcal{R}$ is an equivalence, that is, $\mathcal{R}$ is 
reflexive, symmetric and transitive.
\end{theorem}
The proof  is detailed in Appendix \ref{app-WFH-equiv}, it follows a similar pattern as the proof that strong FH-bisimulation is an equivalence, but technical details are different, and in practice we rely on a variant of the definition of weak FH-bisimilarity; this equivalent version simulates a \emph{weak} open transition with a set of weak open transition. The careful use of the best definition of weak FH-bisimilarity makes the proof similar to the strong FH-bisimulation case.

@@

\begin{theorem}[Congruence for  weak bisimulation]\label{weak-thm-congr-eq}
	Consider an open pNet $P$
 that cannot observe silent actions, of the form 	$\pNet = \mylangle \pNet_i^{i\in I}, \Sort_j^{j\in J}, 
	\set{\symb{SV}}\myrangle$.
	Let $j_0\in J$ be a hole. Let $\pNetQ$ and $Q'$ be two weak FH-bisimilar pNets such that\footnote{Note that $\Sortop(\pNetQ)=\Sortop(\pNetQ')$ is 
	ensured by 
	weak bisimilarity.} 
	$\Sortop(\pNetQ)=\Sortop(\pNetQ')\subseteq\Sort_{j_0}$. Then 
	$\pNet[\pNetQ]_{j_0}$ and 
	$\pNet[\pNetQ']_{j_0}$ are weak FH-bisimilar.
\end{theorem}

\begin{theorem}[Context equivalence for  weak bisimulation]\label{weak-thm-ctxt-eq}
	Consider two  open pNets
	$\pNet = \mylangle \pNet_i^{i\in I}, \Sort_j^{j\in J}, 
	\set{\symb{SV}}\myrangle$ and 	$\pNet' = \mylangle {\pNet'}_i^{i\in I}, 
	\Sort_j^{j\in 
	J}, 	\set{\symb{SV'}}\myrangle$ that are weak FH-bisimilar
	(recall they must have the same holes to be bisimilar) and that cannot observe silent actions.
	Let $j_0\in J$ be a hole, and $\pNetQ$ be a pNet such that $\Sortop(\pNetQ)\subseteq\Sort_{j_0}$. Then 
	$\pNet[\pNetQ]_{j_0}$ and 
	$\pNet'[\pNetQ]_{j_0}$ are weak FH-bisimilar.
\end{theorem}

@@

\begin{theorem}[Composability of weak bisimulation]\label{weak-compos}
	Consider two weak FH-bisimilar pNets with an arbitrary number of holes, such that the two pNets cannot observe silent actions. When replacing, 
	inside those two original pNets, a subset of the holes by weak FH-bisimilar pNets, we 
	obtain two weak FH-bisimilar pNets.
\end{theorem}

@@

\medskip
\noindent
Preliminary remarks:
  \begin{itemize}
    \item Both pNets trivialy verify the ``non-observability''
      condition: the only vectors having $\tau$ as an action of a
      sub-net are of the form ``$< -, \tau, -> \rightarrow \tau$''.
    \item We must take care of variable name conflicts: in our example, the variables of the 2 systems already have different names, but the action parameters occurring in the transitions (\texttt{m}, \texttt{msg}, \texttt{ec}) are the same, that is not correct. In the tools, this is managed by the static semantic layer; in the  example, we rename the only conflicting variables $m$ into $m1$ for \symb{SimpleProtocolSpec}, and $m2$ for \symb{SimpleProtocolImpl}.
    \end{itemize}

@@

Research related to the keyword "Symbolic Bisimulation" refer to two very different domains, namely BDD-like techniques for modelling and computing finite-state bisimulations, that are not related to our topic; and symbolic semantics for data-dependant or high-order systems, that are very close in spirit to our approach. In this last area, we can mention
Calder's work \cite{calder2001symbolic}, that  defines a symbolic semantic for full LOTOS, with a symbolic bisimulation over it;
Borgstrom et al., Liu et al, Delaune et al. and Buscemi et al. providing symbolic semantic and equivalence for different variants of pi calculus respectively \cite{borgstrom2004symbolic,delaune2007symbolic,liu2010complete,buscemi2008open}; and more recently
 Feng et al. provide a symbolic bisimulation for quantum processes \cite{feng2014symbolic}.
All the above works, didn't give a complete approach for verification, and the models on which these works based are definitely different from ours.

@@

\bibliographystyle{lncs/splncs}

@@

              \begin{proof}

       	It is trivial to check reflexivity and symmetry. Here we focus on the
       	transitivity. 
       	To prove transitivity of strong FH-bisimulation on pNets it is sufficient to 
       	prove 
       	transitivity of the strong FH-bisimulation on states. Consider 3 open automata 
       	$\mathcal{T}_1$, $\mathcal{T}_2$, $\mathcal{T}_3$ and states $s$, $t$, $u$ 
       	in those 
       	automata\footnote{We omit the constraints stating that each $s_x,\,t_x,\,u_x$ is 
       	in the 
       	states of 
       		$\mathcal{T}_1,\,\mathcal{T}_2,\,\mathcal{T}_3$ for the sake of readability}.
       	Suppose we have $\mathcal{R}$ an FH-bisimulation relation between states of 
       	$\mathcal{T}_1$ and of  $\mathcal{T}_2$; members of $\mathcal{R}$ are of the form 
       	$(s,t|\Pred_{s,t})$.
       	Suppose we also  have $\mathcal{R}'$ an FH-bisimulation relation between states 
       	of 
       	$\mathcal{T}_2$ and of  $\mathcal{T}_3$; members of $\mathcal{R}'$ are of the 
       	form 
       	$(t,u|\Pred_{t,u})$.

@@

       \end{proof}

@@

   \begin{proof}

Consider rule \TrDeux\ in 
       Definition~\ref{def:operationalSemantics}, applied to the pNet $P[Q]_{j_0}$. 	
	\noindent
\begin{small}	
\begin{mathpar}
    \mprset {vskip=.8ex}
\inferrule
    {
\Leaves(\mylangle {\pNet}_m^{m\in I}, \set{\Sort}, \symb{SV}_k^{k\in 
    	K}\myrangle) \!=\! \pLTS_l^{l\in L} \qquad  	
k\!\in\! K \qquad SV_k \!=\! \SV{(\alpha'_m)^{m \in I_1\uplus I_2\uplus J}}{\alpha'}{e_b} 
\\
\\     	
	\forall m\!\in\! I_1. {\pNet_m 
	\models\openrule
    	{
    	\beta_{j}^{j\in J_m}, \Pred_m, \Post_m}
    	{\ostate{s_{i}^{i \in L_m}} \OTarrow {\alpha_m}
    		\ostate{(s_i^\prime)^{i\in L_m}}} }	
  \qquad
\forall m\!\in\! I_2.		{ \pNet_m 
    	 \models
    	\openrule
    	{\emptyset, \Pred_m, \Post_m}
    	{\ostate{s_m} \OTarrow {\alpha_m}
    		\ostate{s_m'}} }\\\\
     J' = \biguplus_{m\in I_1}\!\! J_m \uplus J 	\\
    	\Pred = \bigwedge_{m\in I_1\uplus I_2}\!\! \Pred_m \land
    	\Predsv(SV_k,\alpha_m^{m\in I_1\uplus I_2},\beta_j^{j\in J},\alpha)\\ 
    	\forall i\in	L\backslash \left(\biguplus_{m\in I_1}\!\! L_m \uplus I_2\right).\,s'_i=s_i \\
    \fresh(\alpha'_m,\alpha',\beta_j,\alpha) 
    }
    {\mylangle {\pNet}_m^{m\in I}, \set{\Sort}, \symb{SV}_k^{k\in K}\myrangle
    	\models
    	{\openrule
    		{
    		{\beta_j}^{j\in J^\prime}, \Pred,  \biguplus_{m\in I_1\uplus I_2} 
    		\Post_m}
    		{\ostate{s_i^{i\in L}} \OTarrow {\alpha}
    			\ostate{(s_i^\prime)^{i\in L}}}
    	}
    }\quad\TrDeux
\end{mathpar}

@@

\begin{proof} 

@@

   This provides the desired conclusion.     \end{proof}

@@

\end{lemma}

>>>>

\subsubsection*{Structure}
This article is organised as follows. Section~\ref{sec:notations}
provides the definition of pNets and introduces the notations used in
this paper, including the definition of open pNets. 
Section~\ref{sec:OT} defines open automata, i.e. automata
with parameters and transitions conditioned by the behaviour of
``holes''; a strong bisimulation equivalence for open automata is also
presented in this section. 
Section~\ref{section:op-semantics} gives
the semantics of open pNets expressed as open automata, and states
compositional properties on the strong bisimulation for open
pNets. 
Section~\ref{sec:weak} defines a weak bisimulation
equivalence on open automata and derives weak bisimilarity for pNets,
together with properties on compositionality of weak bisimulation for
open pNets. 
Finally, Section~\ref{sec:RW} discusses related works and Section~\ref{section:conclusion} concludes the
paper.

@@

The informal semantics of pNets is as follows. pLTSs behave more or less like  classical automata with conditional branching and variables. The actions on the \pLTS s can send or receive values, potentially modifying the value of variables. 
pNets are synchronisation entities: a pNet node composes several sub-pNets and  define how the sub-pNets interact, where a sub-pNet is either a pNet or a pLTS. The synchronisation between sub-pNets is defined by synchronisation vectors (originally introduced by \cite{Arnold1982}) that express how an action of a sub-pNet can be synchronised with actions of other sub-pNet, and how the resulting synchronised action is visible from outside of the pNet. The synchronisation mechanism is very expressive, including pattern-matching/unification between the parameterized actions within the vector, and an additional predicate over their variables.
Consider a pNet node that assembles several pLTSs, the synchronisation vectors specify the way that transitions of the composed pNet are built from the transitions of the sub-nets. This can be seen as "conditional transitions" of the pNet, or alternatively, as a syntax to encode structural operational semantics (SOS rules) of the system: each vector expresses not only the actions emitted by the pNet but also what transitions of the composed pLTSs must occur to trigger this global transition.
Synchronisation vectors can also express the exportation of an action of a sub-pNet to the next level, or to hide an interaction and make it non-observable. Finally, a pNet can leave sub-pNets undefined and instead declare holes with a well-defined signature. Holes can then be filled with a sub-pNet. This is defined as follows.

@@

Figure \ref{SimpleProt:Spec} (right) shows a graphical view of the pNet $\symb{SimpleProtocolSpec}$ that specifies the system. The pNet is made of the composition of two pNets: a $\symb{SimpleSystem}$ node, and a $\symb{PerfectBuffer}$ sub-pNet. 
The full system implementation should be equivalent (e.g.  weakly bisimilar)
to this $\symb{SimpleProtocolSpec}$.  
The pNet has a tree-like structure.  The root node of the tree {\it SimpleSystem} is the top level of the pNet structure. It acts as the parallel operator. It consists of three nodes: two holes $P$ and $Q$ and one sub-pNet, denoted {\it PerfectBuffer}. Nodes of the tree are synchronised using four synchronisation vectors, that  express the possible synchronisations between the parameterised actions of a subset of the nodes. For instance, in the vector $< \texttt{p-send(m)},\texttt{in(m)},\_> \rightarrow \texttt{\nounderline{in(m)}}$ only $P$ and {\it PerfectBuffer} nodes are involved in the synchronisation. The synchronisation between these processes occurs when  process $P$  performs $\texttt{p-send(m)}$  action sending a message, and  the {\it PerfectBuffer} accepts the message through an $\texttt{in(m)}$ action  at the same time; the result that will be returned at upper level is the action  $\texttt{\nounderline{in(m)}}$.

@@

\begin{thm}[FH-Bisimulation is an equivalence]\label{thm-equiv} Suppose $\mathcal{R}$ 
is an FH-bisimulation. Then $\mathcal{R}$ is an equivalence, that is, $\mathcal{R}$ is 
reflexive, symmetric and transitive.
\end{thm}

@@

\begin{thm}[Decidability of FH-bisimulation]
Let $A_1$ and $A_2$ be finite open automata
and $\mathcal{R}$ a relation over their states $\mathcal{S}_1$ and
$\mathcal{S}_2$ constrained by a set of predicates. Assume that
the predicates inclusion is decidable over  
the action algebra $\AlgA$. Then it is decidable whether the relation 
$\mathcal{R}$ is an FH-bisimulation.

\end{thm}

@@

We start with a decomposition property: from one open transition of $P[Q]_{j_0}$, we exhibit 
corresponding behaviours of $P$ and $Q$, and determine the relation between their 
predicates.
\begin{lem}[Open transition decomposition\label{lem-decompose}] Consider two pNets $P$ and $Q$ that are not pLTSs\footnote{A similar lemma can be proven for a pLTS $Q$}.
	Let $\Leaves(Q)=p_l^{l\in L_Q}$ and suppose:
	\[ P[Q]_{j_0}  
		\models
		{\openrule
			{
				\beta_j^{j\in J}, \Pred,  
				\Post}
			{\ostate{s_i^{i\in L}} \OTarrow {\alpha}
				\ostate{s_i'^{\, i\in L}}}
		}
	\]
		with  $J\cap\Holes(Q)\neq\emptyset$ or $\exists i\in L_Q.\,s_i\neq s'_i$, i.e. $Q$ takes part in the reduction.
		 Then there exist $\alpha_Q$, $\Pred\,'$, $\Pred\,''$, 
		$\Post\,'$, $\Post\,''$ s.t.:\\[-2ex]
		%\[
		\begin{mathpar}
		P\models{\openrule
			{
				\beta_j^{j\in (J\setminus \Holes(Q)) \cup \{j_0\}}, 
				\Pred\,',  
				\Post\,'}
			{\ostate{s_i^{i\in L\setminus L_Q}} \OTarrow {\alpha}
				\ostate{s_i'^{\,i\in L\setminus L_Q}}}
		}%\]
	\vspace{-2.2ex}\\\text{and~\qquad\qquad~}
		%\[
		Q\models{\openrule
			{
				\beta_j^{j\in J\cap\Holes(Q)}, \Pred\,'',  
				\Post\,''}
			{\ostate{s_i^{i\in L_Q}} \OTarrow {\alpha_Q}
				\ostate{s_i'^{\,i\in L_Q}}}
		}%\]
		\end{mathpar}
		and  $\Pred \iff \Pred\,'
		\land \Pred\,''\land \alpha_Q=\beta_{j_0}$, $\Post=\Post\,'\uplus 
		\Post\,''$ where $\Post\,''$ is the restriction of $\Post$ over variables of 
		$Q$.
\end{lem}

@@

\begin{lem}[Open transition composition]\label{lem-compose} 
	Suppose $j_0\in J$ and:\\[-1ex]
\begin{mathpar}
%\[
P\models{\openrule
	{
		\beta_j^{j\in J}, 
		\Pred,  
		\Post}
	{\ostate{s_i^{i\in L}} \OTarrow {\alpha}
		\ostate{s_i'^{\, i\in L}}}
}%\]
\quad\text{~~and~~}\quad
%\[
Q\models{\openrule
	{
		\beta_j^{j\in J_Q},
		 \Pred\,',  
		\Post\,'}
	{\ostate{s_i^{i\in L_Q}} \OTarrow {\alpha_Q}
		\ostate{s_i'^{\, i\in L_Q}}}
}%\]
\end{mathpar}
Then, we have:\\[-1ex]
	\[ P[Q]_{j_0}  
	\models
	{\openrule
		{
			\beta_j^{(j\in J\setminus\{j_0\}) \uplus J_Q}, 
			\Pred\land\Pred\,'\land \alpha_Q=\beta_{j_0},  
			\Post\uplus \Post\,'}
		{\ostate{s_i^{i\in L\uplus L_Q}} \OTarrow {\alpha}
			\ostate{s_i'^{\, i\in L\uplus L_Q}}}
	}
	\]
\end{lem}

@@

\begin{thm}[Congruence]\label{thm-congr-eq}
	Consider an open pNet:
	$\pNet = \mylangle \pNet_i^{i\in I}, \Sort_j^{j\in J}, 
	\set{\symb{SV}}\myrangle$.
	Let $j_0\in J$ be a hole. Let $\pNetQ$ and $\pNetQ'$ be two FH-bisimilar pNets such that\footnote{Note that $\Sortop(\pNetQ)=\Sortop(\pNetQ')$ is 
	ensured by 
	strong bisimilarity.} 
	$\Sortop(\pNetQ)=\Sortop(\pNetQ')=\Sort_{j_0}$. Then 
	$\pNet[\pNetQ]_{j_0}$ and 
	$\pNet[\pNetQ']_{j_0}$ are FH-bisimilar.
\end{thm}

\begin{thm}[Context equivalence]\label{thm-ctxt-eq}
	Consider two  open pNets
	$\pNet = \mylangle \pNet_i^{i\in I}, \Sort_j^{j\in J}, 
	\set{\symb{SV}}\myrangle$ and 	$\pNet' = \mylangle {\pNet'}_i^{i\in I}, 
	\Sort_j^{j\in 
	J}, 	\set{\symb{SV'}}\myrangle$ that are FH-bisimilar
	(recall they must have the same holes to be bisimilar).
	Let $j_0\in J$ be a hole, and $Q$ be a pNet such that $\Sortop(Q)=\Sort_{j_0}$. Then 
	$\pNet[Q]_{j_0}$ and 
	$\pNet'[Q]_{j_0}$ are FH-bisimilar.
\end{thm}

Finally, the previous theorems can be composed to state a general theorem about 
composability and FH-bisimilarity.
\begin{thm}[Composability] \label{thm-composability}
	Consider two FH-bisimilar pNets with an arbitrary number of holes, when replacing, 
	inside those two original pNets, a subset of the holes by FH-bisimilar pNets, we 
	obtain two FH-bisimilar pNets.
\end{thm}
This theorem is quite powerful. It somehow implies that the theory of open pNets is convenient to study properties of process composition. Open pNets can indeed be used to study process operators and process algebras, as shown in~\cite{henrio:Forte2016} where compositional properties are extremely useful. In the case of interaction protocols~\cite{BHHM:FACS11}, composition of bisimulation can justify abstractions used in some parts of the application.

@@

\begin{thm}[Weak FH-Bisimulation is an equivalence]\label{thm-weak-equiv} Suppose $\mathcal{R}$ 
is a weak FH-bisimulation. Then $\mathcal{R}$ is an equivalence, that is, $\mathcal{R}$ is 
reflexive, symmetric and transitive.
\end{thm}
The proof  is detailed in Appendix \ref{app-WFH-equiv}, it follows a similar pattern as the proof that strong FH-bisimulation is an equivalence, but technical details are different, and in practice we rely on a variant of the definition of weak FH-bisimilarity; this equivalent version simulates a \emph{weak} open transition with a set of weak open transition. The careful use of the best definition of weak FH-bisimilarity makes the proof similar to the strong FH-bisimulation case.

@@

\begin{thm}[Congruence for  weak bisimulation]\label{weak-thm-congr-eq}
	Consider an open pNet $P$
 that cannot observe silent actions, of the form 	$\pNet = \mylangle \pNet_i^{i\in I}, \Sort_j^{j\in J}, 
	\set{\symb{SV}}\myrangle$.
	Let $j_0\in J$ be a hole. Let $\pNetQ$ and $Q'$ be two weak FH-bisimilar pNets such that\footnote{Note that $\Sortop(\pNetQ)=\Sortop(\pNetQ')$ is 
	ensured by 
	weak bisimilarity.} 
	$\Sortop(\pNetQ)=\Sortop(\pNetQ')\subseteq\Sort_{j_0}$. Then 
	$\pNet[\pNetQ]_{j_0}$ and 
	$\pNet[\pNetQ']_{j_0}$ are weak FH-bisimilar.
\end{thm}

\begin{thm}[Context equivalence for  weak bisimulation]\label{weak-thm-ctxt-eq}
	Consider two  open pNets
	$\pNet = \mylangle \pNet_i^{i\in I}, \Sort_j^{j\in J}, 
	\set{\symb{SV}}\myrangle$ and 	$\pNet' = \mylangle {\pNet'}_i^{i\in I}, 
	\Sort_j^{j\in 
	J}, 	\set{\symb{SV'}}\myrangle$ that are weak FH-bisimilar
	(recall they must have the same holes to be bisimilar) and that cannot observe silent actions.
	Let $j_0\in J$ be a hole, and $\pNetQ$ be a pNet such that $\Sortop(\pNetQ)\subseteq\Sort_{j_0}$. Then 
	$\pNet[\pNetQ]_{j_0}$ and 
	$\pNet'[\pNetQ]_{j_0}$ are weak FH-bisimilar.
\end{thm}

@@

\begin{thm}[Composability of weak bisimulation]\label{weak-compos}
	Consider two weak FH-bisimilar pNets with an arbitrary number of holes, such that the two pNets cannot observe silent actions. When replacing, 
	inside those two original pNets, a subset of the holes by weak FH-bisimilar pNets, we 
	obtain two weak FH-bisimilar pNets.
\end{thm}

@@

\medskip
\noindent
Preliminary remarks:
  \begin{itemize}
    \item Both pNets trivially verify the ``non-observability''
      condition: the only vectors having $\tau$ as an action of a
      sub-net are of the form ``$< -, \tau, -> \rightarrow \tau$''.
    \item We must take care of variable name conflicts: in our example, the variables of the 2 systems already have different names, but the action parameters occurring in the transitions (\texttt{m}, \texttt{msg}, \texttt{ec}) are the same, that is not correct. In the tools, this is managed by the static semantic layer; in the  example, we rename the only conflicting variables $m$ into $m1$ for \symb{SimpleProtocolSpec}, and $m2$ for \symb{SimpleProtocolImpl}.
    \end{itemize}

@@

Research related to the keyword "Symbolic Bisimulation" refer to two very different domains, namely BDD-like techniques for modelling and computing finite-state bisimulations, that are not related to our topic; and symbolic semantics for data-dependant or high-order systems, that are very close in spirit to our approach. In this last area, we can mention
Calder's work \cite{calder2001symbolic}, that  defines a symbolic semantic for full Lotos, with a symbolic bisimulation over it;
Borgstrom et al., Liu et al, Delaune et al. and Buscemi et al. providing symbolic semantic and equivalence for different variants of pi calculus respectively \cite{borgstrom2004symbolic,delaune2007symbolic,liu2010complete,buscemi2008open}; and more recently
 Feng et al. provide a symbolic bisimulation for quantum processes \cite{feng2014symbolic}.
All the above works, did not give a complete approach for verification, and the models on which these works based are definitely different from ours.

@@

\bibliographystyle{alpha}

@@

       	\proof
       	It is trivial to check reflexivity and symmetry. Here we focus on the
       	transitivity. 
       	To prove transitivity of strong FH-bisimulation on pNets it is sufficient to 
       	prove 
       	transitivity of the strong FH-bisimulation on states. Consider 3 open automata 
       	$\mathcal{T}_1$, $\mathcal{T}_2$, $\mathcal{T}_3$ and states $s$, $t$, $u$ 
       	in those 
       	automata\footnote{We omit the constraints stating that each $s_x,\,t_x,\,u_x$ is 
       	in the 
       	states of 
       		$\mathcal{T}_1,\,\mathcal{T}_2,\,\mathcal{T}_3$ for the sake of readability}.
       	Suppose we have $\mathcal{R}$ an FH-bisimulation relation between states of 
       	$\mathcal{T}_1$ and of  $\mathcal{T}_2$; members of $\mathcal{R}$ are of the form 
       	$(s,t|\Pred_{s,t})$.
       	Suppose we also  have $\mathcal{R}'$ an FH-bisimulation relation between states 
       	of 
       	$\mathcal{T}_2$ and of  $\mathcal{T}_3$; members of $\mathcal{R}'$ are of the 
       	form 
       	$(t,u|\Pred_{t,u})$.

@@

%       \end{proof}
\qed

@@

%   \begin{proof}
\proof
Consider rule \TrDeux\ in 
       Definition~\ref{def:operationalSemantics}, applied to the pNet $P[Q]_{j_0}$. 	
	\noindent
\begin{small}	
\begin{mathpar}
    \mprset {vskip=.8ex}
\inferrule
    {
\Leaves(\mylangle {\pNet}_m^{m\in I}, \set{\Sort}, \symb{SV}_k^{k\in 
    	K}\myrangle) \!=\! \pLTS_l^{l\in L} \qquad  	
k\!\in\! K \qquad SV_k \!=\! \SV{(\alpha'_m)^{m \in I_1\uplus I_2\uplus J}}{\alpha'}{e_b} 
\\
\\     	
	\forall m\!\in\! I_1. {\pNet_m 
	\models\openrule
    	{
    	\beta_{j}^{j\in J_m}, \Pred_m, \Post_m}
    	{\ostate{s_{i}^{i \in L_m}} \OTarrow {\alpha_m}
    		\ostate{(s_i^\prime)^{i\in L_m}}} }	
  \qquad
\forall m\!\in\! I_2.		{ \pNet_m 
    	 \models
    	\openrule
    	{\emptyset, \Pred_m, \Post_m}
    	{\ostate{s_m} \OTarrow {\alpha_m}
    		\ostate{s_m'}} }\\\\
     J' = \biguplus_{m\in I_1}\!\! J_m \uplus J 	\\
    	\Pred = \bigwedge_{m\in I_1\uplus I_2}\!\! \Pred_m \land
    	\Predsv(SV_k,\alpha_m^{m\in I_1\uplus I_2},\beta_j^{j\in J},\alpha)\\ 
    	\forall i\in	L\backslash \left(\biguplus_{m\in I_1}\!\! L_m \uplus I_2\right).\,s'_i=s_i \\
    \fresh(\alpha'_m,\alpha',\beta_j,\alpha) 
    }
    {\mylangle {\pNet}_m^{m\in I}, \set{\Sort}, \symb{SV}_k^{k\in K}\myrangle
    	\models
    	{\openrule
    		{
    		{\beta_j}^{j\in J^\prime}, \Pred,  \biguplus_{m\in I_1\uplus I_2} 
    		\Post_m}
    		{\ostate{s_i^{i\in L}} \OTarrow {\alpha}
    			\ostate{(s_i^\prime)^{i\in L}}}
    	}
    }\quad\TrDeux
\end{mathpar}

@@

\proof
%\begin{proof} 

@@

   This provides the desired conclusion.   
\qed

@@

\end{lem}
